---
title: "homework 2"
---

**Due:** Weds ... 11:59 PM. The ideal format for submission is an Rmarkdown or Quarto file, but you can submit it as a Jupyter notebook or as a code file with comments if you prefer. Push your submission to your github repository and send me a note on Piazza to let me know you've done it. You can use R or Python (although the instructions below are R-centric); if you want to use any other language, please check with me first.

**Please make sure your code is reproducible**:

* no absolute path names (e.g. `"C:\My Files\homework"`)
* include any `library()` statements at the top of your file (do not include `install.packages()` statements unless they are commented out)
* make sure I have access to any data files you use in your examples

See also the [R style guide](https://bbolker.github.io/stats720/R_style.html).

If your code isn't reproducible I will return it immediately and ask you to fix it.

1. Analyze the `kyphosis` data (from the `rpart` package). Pick an appropriate family/link function,  Compare the diagnostic plots from base R, from `performance::check_model()`, and from `DHARMa`. Interpret your results and draw coefficient plots.


2. Gopher tortoise example

```{r}

```

https://stackoverflow.com/questions/68686004/how-to-do-negative-binomial-regression-with-the-rms-package-in-r/68703392#68703392

* fit
* fit with `bbmle`
* write your own negative log-likelihood function
3. Complete separation example

```{r}
library(brglm2)
```

https://cran.r-project.org/web/packages/detectseparation/vignettes/separation.html

