---
title: "homework 5"
bibliography: "../stats720.bib"
date: today
date-format: "D MMM YYYY"
---

**Due:** Sunday December 17. The ideal format for submission is an Rmarkdown or Quarto file, but you can submit it as a Jupyter notebook or as a code file with comments if you prefer. Push your submission to your github repository and send me a note on Piazza to let me know you've done it. You can use R or Python (although the instructions below are R-centric); if you want to use any other language, please check with me first.

**Please make sure your code is reproducible**:

* no absolute path names (e.g. `"C:\My Files\homework"`)
* include any `library()` statements at the top of your file (do not include `install.packages()` statements unless they are commented out)
* make sure I have access to any data files you use in your examples

See also the [R style guide](https://bbolker.github.io/stats720/R_style.html).

This assignment will work with the `Contraception` data set you previously explored, but mostly using GAM(M)s (via `mgcv`) rather than GLM(M)s. Throughout, you should focus *only* on the effects of age (scaled/centred for (1) computational stability and (2) comparison of parameter estimates) and urban/rural. In the examples below, I've converting the response variable (`use`) from a factor (Y/N) to a numeric (0/1) variable, again for convenience of downstream processing.

You can use the following summary functions to add a nice summary of the raw data to your plot: **note** you might need to change the name of the `x` and `y` variables to match yours in the `aes()` line below (I decided to create new variables `age_sc` and `use_n` to represent the transformed (centered/numericalized) versions of the `age` and `use` variables; you might have chosen to overwrite them instead).

For plotting, note that `plot(ggeffects::ggemmeans(<model>, terms = c("age [all]", "urban")))` appears to work fairly well for `gam()` output (again with the name of the `age` variable adjusted appropriately).^[(1) the `ggeffects::` part isn't necessary if you've loaded the `ggeffects` package; (2) for reasons I don't yet understand, `ggpredict()` gives weird answers for these models] For some models below you may get an "undefined columns selected" error, but it seems harmless (??)

Whenever I say "quadratic", I mean "quadratic and linear", i.e. don't violate the principle of marginality [@venablesExegeses1998]

Compute binomial mean and CIs: see [this Stack Overflow question](https://stackoverflow.com/questions/77544559/fast-binom-test-in-r-for-multiple-levels-of-confidence/77545378#77545378).

```{r plotfuns}
binom_sum <- function(x, alpha = 0.05) {
    n <- length(x); x <- sum(x)
    est <- x/n
    lwr <- qbeta(alpha/2, x, n-x+1)
    zvals <- !is.na(x) & x == 0
    nvals <- !is.na(x) & x == n
    lwr[zvals] <- 0
    lwr[nvals] <- (alpha/2)^(1/n[nvals])
    upr <- qbeta(1-alpha/2, x+1, n-x)
    upr[zvals] <- 1-(alpha/2)^(1/n[zvals])
    upr[nvals] <- 1
    data.frame(y=est, ymin=lwr, ymax=upr)
}
```

To add points and confidence intervals based on this function to a `ggplot`, add this object (e.g. `plot(ggemmeans(...)) + ssb`)

```{r ssb}
data("Contraception", package = "mlmRev")
Contraception <- transform(Contraception,
                           use_n = as.numeric(use) - 1,
                           age_sc = drop(scale(age)))
ssb <- ggplot2::stat_summary_bin(data = Contraception,
                            fun.data = binom_sum,
                            bins = 10,
                            geom = "pointrange",
                            ## displace urban/rural bars slightly
                            position = ggplot2::position_dodge(width = 0.05),
                            mapping = ggplot2::aes(x = age_sc, y = use_n, colour = urban),
                            fill = NA)
```

These packages could be useful:
```{r setup, message = FALSE}
library(mgcv)
library(lme4)
library(broom)
library(broom.mixed)
library(dotwhisker)
library(ggeffects)
```

1. Fit a model including the effects of age, urban/rural, and their interaction, with the intercept varying among districts, using both `glmer` and `mgcv::gam()` (hint/reminder: `s(..., bs="re")` fits an IID latent variable, i.e. the equivalent of a random intercept, in `gam()`).  Compare the estimates and standard errors for the fixed-effect, non-smooth, non-intercept coefficients (i.e. age, urban/rural, age × urban/rural).

2. Now fit models (using `gam()` only) with (a) a fixed *quadratic* function of age [additive, i.e. `urban` is also in the model but does not interact with age / the age-pattern is assumed to be independent of urban vs. rural] [specifying a quadratic term as `poly(age_sc, 2)` will probably work better than `age_sc + I(age_sc^2)`, although both *should* work]; (b) an effect of age modeled as a thin-plate spline (`bs = "tp"`; this is the default for `s()`). Plot the predictions for each model (again including the binned proportions/binomial CIs for the raw data) and compare them. (Both models should still include a random intercept with respect to `district`.) (For extra credit, figure out how to plot the predictions for both models on the same plot.)

3. Fit models (with the same general parameters as in the previous question, e.g. you should still include a random intercept term) with (a) a quadratic-age $\times$ urban/rural interaction; (b) separate thin-plate splines for `age_sc` for urban and rural settings (use the `by=` argument to `s()`). Do the plots again.

4. Finally, fit (a) a `glmer()` model that includes a fixed quadratic-age by urban/rural interaction and a random effect that allows the quadratic effect of age to vary across districts (i.e., include a `(<quadratic age model>|district)` term) and (b) a `gam()` model that includes different population-level smooths for urban vs rural (using the `by=` specification from the previous model) plus different age-smooths for each district (using a *factor smooth* basis, i.e. `s(age_sc, district, bs = "fs")`).^[This model could take a while to fit. I used `control = gam.control(trace = TRUE, nthreads = 10)` to (1) print lots of output so I could see that something was happening; (2) parallelize the computation. You should restrict the number of threads to less than the number of cores on your computer (`parallel::detectCores()`). On a relatively fast computer with 10 threads, fitting this model took about a minute.] Generate predictions using `predict()` (**not** `ggpredict` etc.) with the following data frame as `newdata`:

```{r newdata}
nd <- with(Contraception,
           expand.grid(urban = levels(urban),
                       age_sc = unique(age_sc),
                       district = unique(district)))
```

Plot the results, for both models, compare. (**not** using `ggpredict()` unless you can figure out how; I used `ggplot2` with `aes(group = interaction(district, urban))`  in `geom_line()`; you could probably figure out how to reshape the data and use `matplot()` ...).

